The negative external effects of deforestation on downstream populations in Brazil

### 2024-08-19 output/documents/thesis_paper/paper.qmd
Final changes to paper before submission; Restructured some code files

---
title: "The negative external effects of deforestation on downstream populations in Brazil"
subtitle: "Master Thesis"
author: "Felix Schulz"
format:
  pdf:
    papersize: a4
    keep-tex: true
    toc: false
    number-sections: true
    link-citations: true
    cite-method: biblatex
    biblatexoptions: style=apa
    include-in-header:  
      - text: |
          \usepackage{graphicx}
          \usepackage{amsfonts}
          \usepackage{amsmath}
          \usepackage{pifont}
          \usepackage{booktabs}
          \usepackage{dcolumn}
          \usepackage{makecell}
          \usepackage{threeparttable}
bibliography: "./Master Thesis.bib"
---

@@

The rainforests of the Amazon are under severe pressure from deforestation. In the two centuries leading up to 2022, 8.6% of primary forests in Brazil were cleared [@worldresourcesinstituteGlobalForestWatch2023]. The importance of these areas is manifold. Their extreme levels of biodiversity have high intrinsic value to the world's ecosystem  [@maslinNewViewsOld2005; @laurila-pantHowValueBiodiversity2015] and local communities [@omettoAmazonDeforestationBrazil2011]. Land use change in primary forests releases large amounts of greenhouse gases, which in turn have negative impacts on a global scale [@malhiClimateChangeDeforestation2008; @harrisGlobalMapsTwentyfirst2021]. As a carbon sink, rainforests bind greenhouse gas, working against climate change [@harrisGlobalMapsTwentyfirst2021]. Logging has lead to an estimated emission of around 36 gigatonnes of CO2-equivalent greenhouse gas emissions from 2001 to 2022 in Brazil alone [@worldresourcesinstituteGlobalForestWatch2023].

Economists model deforestation decisions as a tradeoff between its cost and benefits. We observe that cleared land is reused for agriculture, lifestock or mining [@baronaRolePastureSoybean2010; @sonterMiningDrivesExtensive2017]. These endeavors of alternative use are profitable for their owner but occur costs reaching far beyond the limits of the given areas. Based on the discussed negative impacts on climate change alone, researchers conclude that the social costs through climate change would, if considered by the market, render all agricultural conversion in the Amazon inefficient [@souza-rodriguesDeforestationAmazonUnified2019].

In addition to the global reaching effects of deforestation on the climate, there is a substantial body of literature discussing its impact on local utility in the form of health outcomes. Empirical evidence from medical and geoscientific research find negative effects of deforestation-related wildfires and on malaria prevalence [see for example @reddingtonAirQualityHuman2015; @tuckerlimaDoesDeforestationPromote2017]. Given the existing evidence it appears likely that these dangers to the local communities are not considered to the extent it would be socially desireable [@balboniEconomicsTropicalDeforestation2023].

In the research proposed in this essay I seek to shed a light on the importance of rivers in propagating pollution of deforestation on a vast geographical scale. In other words, I want to find evidence of the role of rivers on health outcomes when there occurs deforestation upstream. I seek to test the hypothesis that deforestation leads to river pollution, impacting areas far from the actual forest in question. Previous work has shown the important role of forests in preventing soil erosion [see for example @wilkinsonPresentSoilErosion1994; @veldkampDeforestationReforestationImpacts2020]. Absent this, one would expect increased river pollution downstream of areas subject to logging. Agricultural land use of cleared land may further increase this effect through livestock-related ground pollution [@montgomerySoilErosionAgricultural2007]. If these concerns turn out substantiated, I will be able to uncover a potentially far-reaching fallout beyond the scope of our current understanding of deforestation's effects.

The impact of deforestation on health outcomes has been subject of several publications [@ellwangerDiversityLossClimate2020; @reddingtonAirQualityHuman2015]. The costs of soil erosion were also discussed in a broad range of literature [see for example @tellesCostsSoilErosion2011]. @balboniEconomicsTropicalDeforestation2023 provide an excellent review on the economics of deforestation. Soil erosion as a driver of pollution and pollution-induced health costs has to the best of knowledge only been covered in a small-scale fashion by @maHumanHealthRisk2014 and therefore remains a gap in the research.

@@

@@

Driven by short-run benefits of lumber or alternative land use, researchers observe unprecedented levels of rainforest degradation in the anthropocene [@malhiClimateChangeDeforestation2008]. This is of elevated importance as any loss of primary forests cannot be undone in any reasonable temporal scope [@shuklaAmazonDeforestationClimate1990; @balboniEconomicsTropicalDeforestation2023]. The scientific evidence for the negative impacts of deforestation raises the concern that these externalities are not internalized by the forest extraction market and that deforestation does not occur at a socially optimal rate. At the core of this thought lies the (monetary) assessment of the societal cost of deforestation against the economical gains from alternative use and timber.

Many economic models have been developed to weigh these benefits and external costs on a local and global scale. In their analysis, @balboniEconomicsTropicalDeforestation2023 describe several such models of decision making for deforestation. Their benchmark model, which they claim to closest resemble the microeconomic process of decisionmaking today, disregards externalities but instead focuses on short-run benefits. In fact, discounting of future income leads to a situation where this benchmark model does not guarantee extraction rates that are at or below a *maximum sustained yield* of forest regrowth. They note that lack of property rights and regulation further increases discounting rates in regions like Brazil.

Extended models then acknowledge the fact that deforestation is associated with social costs that are not internal to the parties with property rights to the forest [@balboniAppendixEconomicsTropical2023]. By modeling this cost as a fixed shift to the private marginal cost, the socially optimal equilibrium lies below the market-achieved private rate. This underlines the necessity to internalize these costs through policy. Economic theory yields several instruments to achieve this goal. Pigouvian taxes on logging or subsidies for landowner to *not* remove a forest are among the popular [@balboniEconomicsTropicalDeforestation2023]. I discuss several such attempts in @sec-policy. It does however remain essential to know the extent of the social costs of deforestation for any such policy to work. My proposed paper therefore has important policy implications by shedding a light on the superlocal health cost of logging across Brazil.

## Health Externalities of Deforestation {#sec-health_effects}

The body of research on impacts of deforestation on human health is manifold. There exists convincing evidence that fires used to clear forests lead to increased air pollution and levels of mortality due to pulmonary diseases [@reddingtonAirQualityHuman2015]. While the latter form of pollution may have lingering effects on health, this mechanism of impact of inherently tied to the moment of first clearing a forest. Other studies focus on latent forms of damage arising from the altered landscape after deforestation. It turns out that several species of mosquito connected to spreading Malaria thrive in the cleared areas with more open land and artificial water infrastructure [for example see @olsonDeforestationMalariaMancio2010]. There is however significant debate if this is in fact the driving mechanisms for some of the observed correlations. Increased human exposure to mosquito habitat, increased immunity and biased data on health outcomes may serve as alternative explanations [@tuckerlimaDoesDeforestationPromote2017].

Research has revealed how soil after deforestation features significantly altered biological properties, reducing its ability to retain nutrient storage essential for agriculture and deteriorating its function as greenhouse gas storage [@veldkampDeforestationReforestationImpacts2020]. Important for the context of this study, reduced erosion resistance is another key finding of many studies [for example see @molinaComplexLandCover2012; @zhengEffectsErosionPatterns2005; and @restrepoa.ImpactoDeforestacionErosion2015]. @veldkampDeforestationReforestationImpacts2020 review how the change in soil properties resulting from deforestation and alternative use decrease the ability of soil to allow vertical water flow, increasing the likelihood of erosion through overland flow. This effect is mediated by the type of soil, the slope of the landscape and the kind of replacing alternative use [@pimentelEnvironmentalEconomicCosts1995]. Ecological accounts of the impact of this process range from negligible [@neillWatershedResponsesAmazon2013] to devastating [@giertzAssessingEffectsLand2005]. @hohnerWildfiresAlterForest2019 describe how wildfires used for deforestation may further amplify erosion.

This now allows me to sketch a clear story of the mechanism I want to survey following the framework of *contamination*, *exposure* and *dose-response* [for example see @graffzivinEnvironmentHealthHuman2013]. Deforestation (the independent variable of the study) impacts soil erosion, leading to river pollution (*contamination*). This water may then be used by the local population (*exposure*), leading to negative health outcomes (*dose-response*). Future statistical analysis may include the discussed mediating factors as controls or interaction terms. This model also allows me to potentially estimate a range of models for every stage of the framework if data is available.

While most research focuses on the on-site productivity losses caused by soil degradation, there is existing evidence on how increased erosion impacts sediment runoff and water pollution off-site. @maHumanHealthRisk2014 describe how deforestation increased toxic contamination in a Chinese region subject to deforestation-induced erosion but only indirectly estimate impacts on health outcomes. Most economic estimations of the externalities are tied to estimating defensive expenditures in water pollution. As early as 1988, @holmesOffsiteImpactSoil1988 finds that this increase in erosion leads to increased defensive water treatment cost. In a more recent study @vincentValuingWaterPurification2016 estimate that forest cover decreases water treatment expenditure in Malaysia.

Extending the context to a more wholistic view on social costs in general, in addition to the health outcomes outlined above, the social costs of soil erosion include dam silation and flooding. @clarkErodingSoilsOfffarm1985 calculated the social cost of offsite erosion at USD 3 billion to USD 13 billion in 1980 and the United States alone. @pataultAnalysisOffsiteEconomic2021 estimate the offsite cost of erosion for two sites in Northern France and find costs in the range of EUR 1,000 to EUR 4,300 per year and kilometer squared.

@@

From an ecological point of view, @malhiClimateChangeDeforestation2008 outline their much-cited recommendation on policy agendas managing deforestation in the Amazon: To avoid reaching catastrophic tipping points, a gross maximum deforestation rate of 30%-40% with limited fragmentation should be targeted. They suggest migration corridors, river conservation and the protection of a core rainforest to the end of preserving biodiversity. Finally, education initiatives should target a reduction in wildfires.

Historically, government development policy in Brazil has been a major driver of deforestation, enabling access through large-scale infrastructure projects and subsidies on lifestock and mining endeavors [@maharGovernmentPoliciesDeforestation1989]. Since 1965, the Forest Code regulates land use in the Amazon [@soares-filhoCrackingBrazilForest2014]. While changes in the 1990s have increased its focus on conservation, conflicting interest of the agricultural sector have limited its requirements and kept compliance low.

Brazil has since made extensive pledges to reduce deforestation-related emissions and implements a range of policy and monitoring instruments to ensure attainment [@tacconiLawEnforcementDeforestation2019]. Since its inception in 2008, the Amazon Fund is important cornerstone policy in Brazil's strategy against deforestation [@correaAmazonFund102019]. It is at its core an implementation of the principles of the Reducing Emissions from Deforestation and Forest Degradation (REDD+) framework. International donations are used to fund conservation projects in monitoring existing regulation, maintaining protected areas, supporting sustainable production and investments in innovations. The central idea is to make payments conditional on environmental efforts and thereby subsidize abatement for those who would not have been able to afford it void the transfers [@balboniEconomicsTropicalDeforestation2023]. The effectiveness and efficiency of this command-and-control approach is however disputed [@correaAmazonFund102019]. *Bolsa Verde*, another example of a REDD+ program, paying poor landowners in the Amazon to maintain forest cover is considered to be cost effective [@RePEc:cgs:wpaper:110].

@@

This thesis is approaches the issue of deforestation, river pollution and health outcomes in 2 stages. First, using water quality sensor data, I estimate if deforestation impacts water quality. Using the insights from this analysis, I then attempt to estimate the impact of deforestation on a range of health outcomes using an IV strategy. In this section I outline my model details and identification strategies.

@@

As discussed above, biochemical, theoretical research suggests that deforestation may facilitate ground erosion. Rainfall may then cause pollutants to be drained into the river system, resulting in pollution. Using a large, unbalanced panel of sensor data, I estimate the correlation between river pollution and deforestation. For each sensor, I identify all upstream rivers and calculate distances from each river segment to the sensor. I estimate panel regression models with sensor- and year-fixed effects:
$$
\text{Pollution}_{it} = \alpha + \beta \cdot \text{Deforestation}_{it} + \gamma_i + \delta_t + \epsilon_{it}
$$

where $\text{Pollution}_{it}$ is the pollution level measured by sensor $i$ in year $t$, $\text{Deforestation}_{it}$ is the measure of deforestation upstream of sensor $i$ in year $t$, $\gamma_i$ represents the sensor-fixed effects, accounting for time-invariant characteristics of each sensor, $\delta_t$ represents the year-fixed effects, accounting for shocks common to all sensors in a given year, and $\epsilon_{it}$ is the error term, which is clustered at the estuary and year level to adjust for within-estuary and -year correlation.

The detail of my data allows me to also investigate how different forms of alternative land use impact river pollution. I estimate an analogous model to the above, regressing sensor measurements on four superclasses of land cover replacement: pasture, agriculture, urban environments and mining. Finally, I investigate how distance interacts with the estimated effects. To this end, I aggregate the data into consecutive distance bins of 50km from the sensor and run a panel fixed effects regression with the following model:

@@

Given the above mechanisms of pollution, if the contaminated water comes into contact with the local population, negative health outcomes are to be expected. Using municipality-level mortality rates, data on hospitalizations and -spending, I attempt to estimate the relationship between deforestation and health. To this end, I regress the respective health indicator on both spatially and temporally cumulative upstream deforestation. The temporal cumulativeness results in an interpretation of the data as *deforestation since 1985*, the first year on record. Spatial cumulation is performed using a spatial weighting matrix closer described in @sec-appendix-hydro. Deforestation in my model is the weighted mean of deforestation in all upstream municipalities with weights decreasing in distance. I include municipality and annual fixed effects and report two-way clustered standard errors.

A key issue in this analysis is the potential endogeneity between deforestation and health outcomes. Endogeneity can arise if there are omitted variables that simultaneously affect both deforestation and health, leading to biased estimates. For example, economic development might influence both the rate of deforestation and healthcare quality, creating a spurious correlation. Additionally, reverse causality could be a concern if poorer health outcomes drive changes in land use, such as increased deforestation to make way for agricultural expansion aimed at improving local food security.

To address these endogeneity concerns, I utilize an instrumental variable (IV) approach. An appropriate instrument is one that is correlated with deforestation but uncorrelated with the error term in the health outcome equation. Since mid-2004 the Detection of Deforestation in Real Time (DETER) system monitors land use changes in the legal Amazon. Satellite imagery is used in an automated system to inform the authorities of deforestation events. Until 2017, daily, 250m-resolution imagery of the Moderate-resolution Imaging Spectroradiometer (MODIS) was used for this analysis. This allowed for the detection of deforestation of areas larger than 25 hectares [@dinizDETERBNewAmazon2015]. Crucially, satellite-based surveillance is impossible if lasting cloud coverage prohibits the view from the sky.

@@

Exogeneity requires the instrument to neither directly (*exogeneity*) nor indirectly influence health outcomes through any other variable than deforestation (*exclusion restriction*). The most obvious concern is that increased cloud coverage comes in hand with weather conditions that have effects on public health. For example, increased rainfall may lead to a rise in waterborne diseases or vector-borne diseases like malaria and dengue fever, which are sensitive to changes in precipitation. The spatially cumulative nature of my outcome variable helps alleviate this concern. Since the independent variable of interest is upstream deforestation, I aggregate deforestation in an area of up to 1000 kilometers in river length. The larger the distance, the less likely weather is going to impact health outcomes at a given municipality. Nevertheless, I do include weather controls, namely average precipitation and temperature in my models.

>>>>

---
format:
  pdf:
    papersize: a4
    fontsize: "12"
    linestretch: "1.25"
    keep-tex: true
    toc: true
    lof: true
    lot: true
    number-sections: true
    link-citations: true
    colorlinks: false
    cite-method: biblatex
    biblatexoptions: style=apa
    include-in-header:  
      - text: |
          \usepackage{graphicx}
          \usepackage{amsfonts}
          \usepackage{amsmath}
          \usepackage{setspace}
          \usepackage{pifont}
          \usepackage{booktabs}
          \usepackage{dcolumn}
          \usepackage{makecell}
          \usepackage{threeparttable}
    include-before-body: 
      - file: "./title_page.tex"
bibliography: "./Master Thesis.bib"
---

@@

The rainforests of the Amazon are under severe pressure from deforestation. In the two centuries leading up to 2022, 8.6% of primary forests in Brazil were cleared [@worldresourcesinstituteGlobalForestWatch2023]. The importance of these areas is manifold. Their extreme levels of biodiversity have high intrinsic value to the world's ecosystem  [@maslinNewViewsOld2005; @laurila-pantHowValueBiodiversity2015] and local communities [@omettoAmazonDeforestationBrazil2011]. Land use change in primary forests releases large amounts of greenhouse gases, which in turn have negative impacts on a global scale [@malhiClimateChangeDeforestation2008; @harrisGlobalMapsTwentyfirst2021]. As a carbon sink, rainforests bind greenhouse gases, working against climate change [@harrisGlobalMapsTwentyfirst2021]. Logging has led to an estimated emission of around 36 gigatonnes of CO2-equivalent greenhouse gas emissions from 2001 to 2022 in Brazil alone [@worldresourcesinstituteGlobalForestWatch2023].

Economists model deforestation decisions as a tradeoff between its cost and benefits. One can observe that cleared land is reused for agriculture, livestock, or mining [@baronaRolePastureSoybean2010; @sonterMiningDrivesExtensive2017]. These endeavors of alternative use are profitable for their owner but also have costs reaching beyond the limits of the given areas. Based on the discussed negative impacts of climate change alone, researchers conclude that the social costs through climate change would, if considered by the market, render all agricultural conversion in the Amazon inefficient [@souza-rodriguesDeforestationAmazonUnified2019].

In addition to the global-reaching effects of deforestation on the climate, there is a substantial body of literature discussing its impact on local utility in the form of health outcomes. Empirical evidence from medical and geoscientific research finds negative effects of deforestation-related wildfires and effects on malaria prevalence [see for example @reddingtonAirQualityHuman2015; @tuckerlimaDoesDeforestationPromote2017]. Given the existing evidence it appears likely that these dangers to the local communities are not considered to the extent it would be socially desirable [@balboniEconomicsTropicalDeforestation2023].

Previous work has shown the important role of forests in preventing soil erosion [see for example @wilkinsonPresentSoilErosion1994; @veldkampDeforestationReforestationImpacts2020]. Absent this, one would expect increased river pollution downstream of areas subject to logging. Agricultural land use of cleared land may further increase this effect through livestock-related ground pollution [@montgomerySoilErosionAgricultural2007]. In this thesis, I seek to shed light on the importance of rivers in propagating pollution of deforestation on a vast geographical scale. In other words, I want to find evidence of the role of waterways on health outcomes when there occurs deforestation upstream. I seek to test the hypothesis that

1. Deforestation leads to river pollution, and that 
2. Deforestation impacts health through the river system.

I attempt to test these hypotheses with a novel combination of vast public data. I aggregate deforestation around rivers and observe changes in health outcomes down the course of the river. To this end, I compute a detailed model of all Brazilian rivers and their corresponding drainage areas. Using sensor-measured information on river pollution I estimate the correlation between river contamination and upstream land use. To test my second hypothesis, I assemble a data set of deforestation and health outcomes in all Brazilian municipalities between 1995 and 2022. Using an instrumental variables approach I aim to isolate the causal relationship between the decline of forests and mortality, hospitalization, and health care expenditure. Potentially exogenous variation in cloud cover serves as my instrument. Because monitoring and enforcement of illegal logging is based on satellite technology, cloud cover restricts or even prohibits effective law enforcement, leading to increased deforestation.

My empirical analysis does not find a clear and significant connection between forest removal, pollution, and health outcomes. On aggregate, there is some evidence that upstream deforestation is associated with increased pH levels and total residue in river water. Mining is shown to have the strongest correlation, mostly with water turbidity and biochemical oxygen demand. The picture is however ambiguous: Unexpectedly I also find decreased biochemical oxygen demand in areas with increased upstream deforestation. Turbidity levels are on average lower when agriculture replaces forests. There is also no conclusive evidence that nitrogen and nitrate levels are altered by farming activity.

The instrumental variable regression focusing on health outcomes also yields only inconclusive results. I do not find statistically significant effects on mortality, infant mortality, and health care expenditure. There is however evidence for an impact on hospitalization. I find that a decrease in forest area upstream of 10 percentage points is associated with an increase of around 2.5 standard deviations in hospitalizations. This result holds when modeling spatial autocorrelation in the error term in a spatial error model. The results are however sensitive to parameter choices and do not hold in OLS or when altering the temporal or spatial scope. I therefore refrain from attaching any causal interpretation to my findings and conclude that my approach cannot confirm the health impacts of deforestation.

The impact of deforestation on health outcomes has been the subject of several publications [@ellwangerDiversityLossClimate2020; @reddingtonAirQualityHuman2015]. The costs of soil erosion were also discussed in a broad range of literature [see for example @tellesCostsSoilErosion2011]. @balboniEconomicsTropicalDeforestation2023 provide an excellent review on the economics of deforestation. Soil erosion as a driver of pollution and pollution-induced health costs has to the best of my knowledge only been covered in a small-scale fashion by @maHumanHealthRisk2014 and therefore remains a gap in the research.

The remainder of my thesis is structured as follows. In a literature review, I first introduce theoretical evidence on the mechanism of impact surveyed. Following that, I introduce my model and provide evidence for my approach toward statistical identification. In the fourth section, I elaborate on the data used. After that, I present the results of my empirical analysis and then conclude with a brief discussion.

@@

For decades, researchers have been observing declines in forested areas across the globe. The reasons for this are complex. Directly observable is that in most cases forests make way for some form of economic exploitation. Farming, pasture, mining, and urbanization are frequently observed on previously cleared land [@fearnside1987causes; @chakravarty2012deforestation; @austinWhatCausesDeforestation2019]. Underlying factors differ between the economic and institutional contexts of countries. In the case of Brazil, the process is facilitated by infrastructure projects, tax incentives, and the weakness of institutions [@fearnside1987causes; @binswangerBrazilianPoliciesThat1991; @araujoPropertyRightsDeforestation2009]. In this literature review, I highlight the effects of deforestation, how they are (not) represented in economic evaluations and finally introduce the legal framework governing Brazilian forests.

## Effects of Deforestation {#sec-health_effects}

Deforestation is driven by local actors who financially benefit from woodland removal. Timber extraction, cattle farming and soybean production on the cleared land have considerably higher short-run individual payoffs than rents gained from actions maintaining the state of vegetation [@fearnside1987causes; @seymourReducingTropicalDeforestation2019]. In addition, weak property rights have been shown to lead to land grabbing and more importantly attempts to prevent the latter by establishing claims by clearing the land [@fearnside1987causes; @araujoPropertyRightsDeforestation2009]. Large-scale dam projects with strong impacts on forests are a cornerstone of the Brazilian strategy to secure sustainable energy production [@kahnFalseShadesGreen2014]. The resulting agricultural landscape is however perceived as highly suboptimal. For a variety of reasons, agricultural efficiency in deforested areas in the Amazon is low [@nobreLanduseClimateChange2016; @ferreiraFarmSizeLand2020]. Therefore, despite the significant repurposing and exploitation of primary forests, economic development in the Amazon has not progressed significantly [@garrettForestsSustainableDevelopment2021]. Given the limited evidence of the economic benefits of logging, it is especially important to also observe its downsides.

The body of research on the environmental harm of deforestation is vast. The release of carbon stored in the woodland systems is a major contributor to global climate change [@houghtonGlobalEffectsTropical1990; @harrisGlobalMapsTwentyfirst2021]. Locally, models predict strong decreases in annual rainfall and concentration of precipitation in extreme events [@werthLocalGlobalEffects2002; @lawrenceEffectsTropicalDeforestation2015; @nobreLanduseClimateChange2016]. Dam projects are perceived as costly to the environment, largely offsetting the environmental upsides of sustainable production [@kahnFalseShadesGreen2014].

In addition to these effects on the climate, researchers also highlight the health costs related to deforestation. There exists convincing evidence that fires used to clear forests lead to increased air pollution and levels of mortality due to pulmonary diseases [@reddingtonAirQualityHuman2015]. Other studies focus on latent forms of damage arising from the altered landscape after deforestation. It turns out that several species of mosquito connected to spreading Malaria thrive in the cleared areas with more open land and artificial water infrastructure [for example see @olsonDeforestationMalariaMancio2010]. There is however significant debate if this is the driving mechanism for some of the observed correlations. Increased human exposure to mosquito habitat, increased immunity and biased data on health outcomes may serve as alternative explanations [@tuckerlimaDoesDeforestationPromote2017].

Research has revealed how soil after deforestation features significantly altered biological properties, reducing its ability to retain nutrient storage essential for agriculture and deteriorating its function as greenhouse gas storage [@veldkampDeforestationReforestationImpacts2020]. Important for the context of this study, reduced erosion resistance is another key finding of many studies [for example see @molinaComplexLandCover2012; @zhengEffectsErosionPatterns2005; and @restrepoa.ImpactoDeforestacionErosion2015]. @veldkampDeforestationReforestationImpacts2020 review how the change in soil properties resulting from deforestation and alternative use decreases the ability of soil to allow vertical water flow, increasing the likelihood of erosion through overland flow. This effect is mediated by the type of soil, the slope of the landscape, and the kind of replacement alternative use [@pimentelEnvironmentalEconomicCosts1995]. Ecological accounts of the impact of this process range from negligible [@neillWatershedResponsesAmazon2013] to devastating [@giertzAssessingEffectsLand2005]. @hohnerWildfiresAlterForest2019 describe how wildfires used for deforestation may further amplify erosion.

This now allows me to sketch a clear story of the mechanism I want to survey following the framework of *contamination*, *exposure*, and *dose-response* [for example see @graffzivinEnvironmentHealthHuman2013]. Deforestation (the independent variable of my study) impacts soil erosion, leading to river pollution (*contamination*). This water may then be used by the local population (*exposure*), leading to negative health outcomes (*dose-response*).

While most research focuses on the on-site productivity losses caused by soil degradation, there is existing evidence on how increased erosion impacts sediment runoff and water pollution off-site. @maHumanHealthRisk2014 describe how deforestation increased toxic contamination in a Chinese region subject to deforestation-induced erosion but only indirectly estimate impacts on health outcomes. Most economic estimations of the externalities are tied to estimating defensive expenditures in wa‚ter pollution. As early as 1988, @holmesOffsiteImpactSoil1988 finds that this increase in erosion leads to increased defensive water treatment costs. In a more recent study @vincentValuingWaterPurification2016 estimate that forest cover decreases water treatment expenditure in Malaysia. Extending the context to a more holistic view of the overall social costs of soil erosion, in addition to the health outcomes outlined above, dam siltation and flooding need to be considered. @clarkErodingSoilsOfffarm1985 calculate the social cost of offsite erosion at USD 3 billion to USD 13 billion in 1980 and the United States alone. @pataultAnalysisOffsiteEconomic2021 estimate the offsite cost of erosion for two sites in Northern France and find costs in the range of EUR 1,000 to EUR 4,300 per year and kilometer squared.

@@

The clearing of forests provides the actors with some clear immediate benefits. On the other hand, I have presented considerable scientific evidence for the negative impacts of deforestation. This raises the concern that these externalities are not internalized by the forest extraction market and that deforestation does not occur at a socially optimal rate. This is of elevated importance as any loss of primary forests cannot be undone in any reasonable temporal scope [@shuklaAmazonDeforestationClimate1990; @balboniEconomicsTropicalDeforestation2023]. At the core of this thought lies the (monetary) assessment of the societal cost of deforestation against the economic gains from alternative use and timber.

Many economic models have been developed to weigh these benefits and external costs on a local and global scale. In their analysis, @balboniEconomicsTropicalDeforestation2023 describe several such models of decision-making for deforestation. Their benchmark model, which they claim to resemble the microeconomic process of decision-making today, disregards externalities but instead focuses on short-run benefits. Discounting of future income leads to a situation where this benchmark model does not guarantee extraction rates that are at or below a *maximum sustained yield* of forest regrowth. They note that the lack of property rights and regulation further increases discounting rates in regions like Brazil.

Extended models then acknowledge the fact that deforestation is associated with social costs that are not internal to the parties with property rights to the forest [@balboniAppendixEconomicsTropical2023]. By modeling this cost as a fixed shift to the private marginal cost, the socially optimal equilibrium lies below the market-achieved private rate. This underlines the necessity to internalize these costs through policy. Economic theory yields several instruments to achieve this goal. Pigouvian taxes on logging or subsidies for landowners to *not* remove a forest are among the most popular in research [@balboniEconomicsTropicalDeforestation2023]. I discuss several such attempts in @sec-policy. It does however remain essential to know the extent of the social costs of deforestation for any such policy to work. My thesis therefore has important policy implications by shedding light on the superlocal health cost of logging across Brazil.

@@

From an ecological point of view, @malhiClimateChangeDeforestation2008 outline their much-cited recommendation on policy agendas managing deforestation in the Amazon: To avoid reaching catastrophic tipping points, a gross maximum deforestation rate of 30%-40% with limited fragmentation should be targeted. They suggest migration corridors, river conservation, and the protection of a core rainforest to the end of preserving biodiversity. Finally, education initiatives should target a reduction in wildfires.

Historically, development policy in Brazil has been a major driver of deforestation, enabling access through large-scale infrastructure projects and subsidies on livestock and mining endeavors [@maharGovernmentPoliciesDeforestation1989]. Since 1965, the Forest Code regulates land use in the Amazon [@soares-filhoCrackingBrazilForest2014]. While changes in the 1990s have increased its focus on conservation, strong political pressure of the agricultural sector has limited its requirements and kept compliance low.

Brazil has since made extensive pledges to reduce deforestation-related emissions and implements a range of policy and monitoring instruments to ensure attainment [@tacconiLawEnforcementDeforestation2019]. Since its inception in 2008, the Amazon Fund has been an important cornerstone policy in Brazil's strategy against deforestation [@correaAmazonFund102019]. It is at its core an implementation of the principles of the Reducing Emissions from Deforestation and Forest Degradation (REDD+) framework. International donations are used to fund conservation projects in monitoring existing regulations, maintaining protected areas, supporting sustainable production, and investing in innovations. The central idea is to make payments conditional on environmental efforts and thereby subsidize abatement for those who would not have been able to afford it without the transfers [@balboniEconomicsTropicalDeforestation2023]. The effectiveness and efficiency of this command-and-control approach are however disputed [@correaAmazonFund102019]. *Bolsa Verde*, another example of a REDD+ program, paying poor landowners in the Amazon to maintain forest cover is considered to be cost-effective [@RePEc:cgs:wpaper:110].

@@

This thesis approaches the issue of deforestation, river pollution, and health outcomes in 2 steps. First, using water quality sensor data, I estimate if deforestation impacts water quality. Using the insights from this analysis, I then attempt to estimate the impact of deforestation on a range of health outcomes using an IV strategy. In this section, I outline my model details and identification strategies.

@@

As discussed above, theoretical biochemical research suggests that deforestation facilitates ground erosion. Rainfall may then cause pollutants to be drained into the river system, resulting in pollution. Using a large, unbalanced panel of sensor data, I estimate the correlation between river pollution and deforestation. For each sensor, I identify all upstream rivers and calculate distances from each river segment to the sensor. I estimate panel regression models with sensor- and year-fixed effects of the form:
$$
\text{Pollution}_{it} = \alpha + \beta \cdot \text{Deforestation}_{it} + \gamma_i + \delta_t + \epsilon_{it}
$$

where $\text{Pollution}_{it}$ is the pollution level measured by sensor $i$ in year $t$, $\text{Deforestation}_{it}$ is the measure of deforestation upstream of sensor $i$ in year $t$, $\gamma_i$ represents the sensor-fixed effects, accounting for time-invariant characteristics of each sensor, $\delta_t$ represents the year-fixed effects, accounting for shocks common to all sensors in a given year, and $\epsilon_{it}$ is the error term.

The detail of my data allows me to also investigate how different forms of alternative land use impact river pollution. I estimate an analogous model to the above, regressing sensor measurements on four superclasses of land cover replacement: pasture, agriculture, urban buildup, and mining. Finally, I investigate how distance interacts with the estimated effects. To this end, I aggregate the data into consecutive distance bins of 50km from the sensor and run a panel fixed effects regression with the following model:

@@

If contaminated water comes into contact with the local population, negative health outcomes are to be expected. Using municipality-level mortality rates, data on hospitalizations, and -spending, I attempt to estimate the relationship between deforestation and health. To this end, I regress the respective health indicators on spatially cumulative upstream forest cover. Spatial cumulation is performed using a spatial weighting matrix closer described in @sec-appendix-hydro. In other words, forest cover in my model is the weighted share of the area covered with forest in all upstream municipalities with weights decreasing in distance. Please note that here I am pivoting away from only looking at year-to-year changes in land cover. This has an effect on the interpretation of the results. Using this approach, I consider the *stock* of land cover and therefore the lasting effect of decreasing forest cover. I understand that this approach may suffer from spurious variation: The decreasing trend in cumulative forest cover (and the increasing trends in other land cover categories) might coincide with trends in unobservables, yielding biased coefficients. I nevertheless prefer this approach as the interpretation of results is closer to the relationship of interest. Finally, I include municipality and annual fixed effects and report two-way clustered standard errors.

A key issue in this analysis is the potential endogeneity of forest cover in regard to health outcomes. Endogeneity can arise if there are omitted variables that simultaneously affect both deforestation and health, leading to biased estimates. For example, economic development might influence both the rate of deforestation and healthcare quality, creating a spurious correlation. Additionally, reverse causality could be a concern if poorer health outcomes drive changes in land use, such as increased deforestation to make way for agricultural expansion aimed at improving local food security.

To address these endogeneity concerns, I utilize an instrumental variable (IV) approach. An appropriate instrument is one that is correlated with deforestation but uncorrelated with the error term in the health outcome equation. Since mid-2004 the Detection of Deforestation in Real Time (DETER) system monitors land use changes in the legal Amazon. Satellite imagery is used in an automated system to inform the authorities of deforestation events. Until 2017, daily, 250m-resolution imagery of the Moderate-resolution Imaging Spectroradiometer (MODIS) was used for this analysis. This allowed for the detection of deforestation of areas larger than 25 hectares [@dinizDETERBNewAmazon2015]. Crucially, satellite-based surveillance is impossible if lasting cloud coverage prohibits a view from the sky.

@@

Exogeneity requires the instrument to neither directly (*exogeneity*) nor indirectly influence health outcomes through any other variable than deforestation (*exclusion restriction*). The most obvious concern is that increased cloud coverage is associated with weather conditions that have effects on public health. For example, increased rainfall may lead to a rise in waterborne diseases or vector-borne diseases like malaria and dengue fever, which are sensitive to changes in precipitation. The spatially cumulative nature of my outcome variable helps alleviate this concern. Since the independent variable of interest is upstream deforestation, I aggregate deforestation in an area of up to 1000 kilometers in river length. Faraway weather is less likely to impact health outcomes in a given municipality. Nevertheless, I do include weather controls, namely average precipitation and temperature in my models.

In addition to these theoretical arguments, I provide a placebo test for the validity of the instrument in @tbl-reg-placebo. Using an alternative data source on cloud cover, I test whether the latter predicts deforestation before DETER was initiated and in areas outside the legal Amazon where it is active. Since DETER only reports cloud cover for the areas and the duration in which it is active, I obtain data on cloud cover from ESA and aggregate them by municipality as discussed above. Importantly, both cloud cover indicators are constructed differently: While DETER reports the share of the municipality in which no observation was possible within a year, ESA utilized the higher temporal detail of their satellite data to calculate the overall share of cloud cover within the municipality and year. To show that both are approximately equivalent, I first regress the DETER on the ESA data and find that the former is explained well by the latter. This allows me to estimate the placebo test using the ESA instead of the DETER data. In the second column, I test whether cloud cover predicts deforestation rates in and outside the legal Amazon before DETER was active. Encouragingly, none of the coefficients are statistically significant. After 2005, when DETER was initiated, cloud cover only predicted deforestation in the legal Amazon. This finding is statistically significant at all conventional levels. The instrument therefore passes my placebo test.
